---
title: 'SciX: Genies'
author: "Lachlan Gray"
date: "19/06/2022"
output: html_document
runtime: shiny
---
### Introduction

This file is an Rmarkdown file. Upon successfully installed R and RStudio you should be able to follow the instructions below. If you run into errors please flag on the GitHub page or search the error in Google.

To run each chunk of code press the green arrow next to the command or select the code and run with 'command + enter'.

Opening this Rmarkdown file will place us into the required working directory. A directory is another name for folders on the computer. All of the plots and files we generate will be saved to the working directory.
We can find which directory we are in with the getwd() command:

```{r echo=FALSE}
getwd()
```

For MacOS and Linux getwd() should return the following path:
"/Users/USER/Desktop/SciX-main/HD"

We can then see which files are available in this directory with the dir() command:
```{r setup, include=FALSE}
dir()
```


### Setting up R
```{r setup, include=FALSE}
install.packages('knitr', verbose=F)
knitr::opts_chunk$set(echo = TRUE)
r <- getOption("repos")
r["CRAN"] <- "https://cran.csiro.au/"
options(repos = r)
```

### Install packages 
If asked to update all/some/none just enter 'a' in the console below.
```{r, echo=FALSE}
install.packages('ggplot2', verbose = F)
install.packages('reshape2', verbose = F)
install.packages('ggrepel', verbose = F)
install.packages('ggupset', verbose = F)
install.packages('BiocManager', verbose = F)
install.packages('statmod', verbose = F)
install.packages('pheatmap', verbose = F)
BiocManager::install("qvalue")
BiocManager::install("edgeR")
BiocManager::install("clusterProfiler")
```

### Load the packages
```{r echo=FALSE}
library(ggplot2, verbose = F)
library(qvalue, verbose = F)
library(reshape2, verbose = F)
library(ggrepel, verbose = F)
library(ggupset, verbose = F)
library(enrichplot, verbose = F)
library(edgeR, verbose = F)
library(DOSE, verbose = F)
library(clusterProfiler, verbose = F)
library(statmod, verbose=F)
library(pheatmap, verbose=F)
```

## Set working directory
We need to tell R which folder (known as directory) our data is located and where we want to store our results.
This directory will be the unzipped SciX-main directory we just downloaded.  
As naming paths are different in Mac and PC please make sure you run the appropriate line of code.

### Read in RNA sequencing count matrix. This is the data that we will be using for this experiment
The read.csv command allows us to load a comma separated file into R. This file contains data in the form of a matrix (a grid of numbers). The "header" option is set to "T" which means that the first row of the file contains the names of the columns. The "row.names" option is set to "1" which means that the first column of the file containing the gene names is used to name each row.
We then print the first five rows of the matrix (which includes all of the columns) to the screen. It also prints the dimensions of the matrix, which tells us the number of rows (genes) and columns (individuals) in the matrix.
```{r echo=FALSE}
# Read in comma separated count matrix
count.matrix <- read.csv('PD.data.csv', 
                         header=T, row.names = 1)
# print the first five rows and all of the columns
count.matrix[1:5,]
# Print the dimensions of this matrix. This tells us we have 58,721 genes and 13 individuals
dim(count.matrix)
```

### Read in sample metadata
```{r echo=FALSE}
metadata <- unique(read.delim('PD.metadata.txt')[,-1])
print(metadata)
```

### To make the column names more informative we replace with metadata$condition column
```{r echo=FALSE}
count.matrix <- count.matrix[,metadata$individual]
colnames(count.matrix) <- c(paste(metadata$condition[1:9], 1:9, sep='_'),
                        paste(metadata$condition[10:16], 1:7, sep='_'))
# Print first 5 genes
count.matrix[1:5,]
```

# Differential expression analysis with edgeR likelihood ratio test
We will perform a statistical test to determine which genes are different between our conditions. For this, we will use the likelihood ratio test which takes models from each condition and compares them.
We then make our disease samples the reference group. This tells us the difference in gene expression in relation to our disease group. For example, a gene with a positive (+) logFC is upregulated in disease and a negative (-) logFC is downregulated in disease.
We then filter out lowly expressed genes, normalise the expression values and perform the test.
To visualise our results, we create plots to show differentially expressed genes.

```{r echo=FALSE}
# Create a dataframe containing all of the required information
targets = data.frame(group = gsub('_[0-9]+', '', 
                    metadata$condition),
                    sex = metadata$sex,
                    age = metadata$age)
# Create the model which considers the sex, age and condition in the analysis
design <- model.matrix(~0+sex+age+group, data=targets)
y = DGEList(counts = count.matrix, group = targets$group)
# Change the reference group to the disease group
y$samples$group <- factor(y$samples$group, levels=c('disease', 'control'))
# Filter lowly expressed genes
keep <- filterByExpr(y)
y <- y[keep, ,keep.lib.sizes=FALSE]
# Normalise the data to remove between sample differences
y <- calcNormFactors(y, method = "TMM")
# Transform data into a counts per million value
tmm <- cpm(y)
# Estimate the dispersion in the data
y <- estimateDisp(y, design, robust=TRUE)
# Fit data to a model
fit <- glmQLFit(y, design)
# Likelihood ratio test to compare models
lrt <- glmLRT(fit)
tt <- topTags(lrt, n=nrow(y), p.value=0.05)
# Smearplot
plotSmear(lrt, de.tags=rownames(tt$table))
pdf('smearplot.pdf')
plotSmear(lrt, de.tags=rownames(tt$table))
dev.off()
print(summary(decideTests(lrt)))
lrt <- as.data.frame(lrt)
lrt$FDR <- qvalue(lrt$PValue)$qvalues
gene <- gsub('ENSG[0-9]+.[0-9]+\\|', '', rownames(lrt))
lrt <- cbind(gene,lrt)
# Top 10 differentially expressed genes
lrt[order(lrt$FDR),][1:10,]

# MDS plot
mds <- plotMDS(y, top=1000, plot=F)
toplot <- data.frame(Dim1 = mds$x, Dim2=mds$y, group=factor(y$samples$group), sex=factor(metadata$sex))
mds.plot <- ggplot(toplot, aes(Dim1, Dim2, colour=group)) + 
  geom_point(aes(shape=sex), size=5) +
  scale_shape_manual(values=c(17, 15)) +
  geom_text_repel(data=toplot, label=rownames(toplot)) +
  ggtitle('Multidimensional scaling (MDS) plot')

pdf('MDS.plot.pdf')
mds.plot
dev.off()
```

### Save result file to working directory
```{r echo=TRUE}
write.table(lrt, row.names = F, sep = "\t", 'edgeR-LRT.PD.txt')
```

## Displaying results in volcano plot
This plot displays the log fold-change and false dicovery rate for each gene. You can select the number of genes to label with the **n.genes** variable below.
```{r echo=FALSE}
n.genes = 20

threshold <- lrt$FDR < 0.05 & abs(lrt$logFC) > 0.5
lrt$threshold <- threshold

# Order on FDR
ordered <- lrt[order(lrt$FDR),]
volcano.plot <- ggplot(lrt) +
  geom_point(aes(x=logFC, y=-log10(FDR), colour=threshold)) +
  geom_text_repel(data=head(ordered, n.genes), aes(x=logFC, y=-log10(FDR), label=gene)) +
  ggtitle("Volcano Plot: Parkinson's Disease") +
  xlab("log2 fold change") +
  ylab("-log10 FDR") +
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25))) +
  scale_color_discrete(name = "DEG")
volcano.plot

pdf(paste0('Volcano.Plot.top', n.genes, '.pdf'))
volcano.plot
dev.off()
```

## Display results as heatmap
This plot will show gene expression in a heatmap where the colour of the tile indicates gene expression in each sample. You can select the number of genes to show with the **n.genes** variable below.

```{r echo=FALSE}
n.genes <- 20
mat <- tmm[head(order(lrt$FDR), n.genes),]
rownames(mat) <- gsub('ENSG[0-9]+.[0-9]+\\|', '', rownames(mat))
pheatmap::pheatmap(scale(mat))

pdf(paste0('Heatmap.top', n.genes, '.pdf'))
pheatmap::pheatmap(scale(mat))
dev.off()
```

# Over Representation Analysis (ORA)
```{r echo=FALSE}
gene.set <- read.gmt('../reactome.gmt')
sigOE <- subset(lrt, abs(logFC) > 0.5 & FDR < 0.05)

ego <- enricher(gene = sigOE$gene, 
         universe = lrt$gene,
         pAdjustMethod = "fdr", 
         qvalueCutoff = 0.01,
         TERM2GENE = gene.set)

dotplot(ego, showCategory=20, orderBy="GeneRatio")
pdf('reactome.ORA.pdf', width = 14)
dotplot(ego, showCategory=20, orderBy="GeneRatio")
dev.off()

upsetplot(ego)
pdf('reactome.upsetplot.pdf')
upsetplot(ego)
dev.off()
```

### Match genes to DisGeneNet and perform chi-squared test
```{r echo=FALSE}
disgene <- read.delim('disgenenet.PD.tsv', header=T, sep="\t")
disgene <- unique(disgene$Gene)

deg <- subset(lrt, abs(logFC) > 0.5 & FDR < 0.05)
not.deg <- subset(lrt, FDR > 0.05)
up <- subset(lrt, logFC > 0.5 & FDR < 0.05)
down <- subset(lrt, logFC < -0.5 & FDR < 0.05)

# calculate numbers for different conditions
a <- lrt[lrt$FDR < 0.05 & abs(lrt$logFC) > 0.5 & lrt$gene %in% disgene,]$gene
b <- lrt[lrt$FDR < 0.05 & abs(lrt$logFC) > 0.5 & !(lrt$gene %in% disgene),]$gene
c <- lrt[lrt$FDR > 0.05 & lrt$gene %in% disgene,]$gene
d <- lrt[lrt$FDR > 0.05 & !(lrt$gene %in% disgene),]$gene
ctable <- matrix(c(length(a),length(b),length(c),length(d)),nrow=2, ncol=2)
chisq <- chisq.test(ctable)

print('Expected values')
chisq$expected

print('Observed values')
chisq$observed

print('Pearson residuals') #Positive values indicate a positive association
chisq$residuals

print('chi.squared p.value')
chisq$p.value

# ORA of significant genes known to cause disease
ego <- enricher(gene = a, 
                universe = lrt$gene,
                pAdjustMethod = "fdr", 
                qvalueCutoff = 0.01,
                TERM2GENE = gene.set)
dotplot(ego, orderBy="GeneRatio", font.size=10, label_format=20)
pdf('disgenenet.ORA.pdf', width=14)
dotplot(ego, orderBy="GeneRatio", font.size=12, label_format=20)
dev.off()
```

### Extract genes from interesting pathway. Select pathway with **pathway** variable
```{r echo=FALSE}
pathway <- 3
plot.genes <- unlist(strsplit(data.frame(ego)[pathway,8], '/'))
gene.name <- 
rownames(tmm) <- gsub('ENSG[0-9]+.[0-9]+\\|', '', rownames(tmm))
colnames(tmm) <- gsub('[0-9]+', '', colnames(tmm))
plot.data <- subset(tmm, rownames(tmm) %in% plot.genes)
plot.data <- plot.data[order(rowSums(plot.data), decreasing = T),]
plot.data <- melt(plot.data)
colnames(plot.data) <- c('gene', 'condition', 'cpm')
boxplot.plot <- ggplot(plot.data, aes(x=gene, y=cpm, fill=condition)) + 
  geom_boxplot() +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  ggtitle(paste('', ego[pathway,1])) +
  ggtitle(paste('Expression of', ego[pathway,1], 'genes'))

boxplot.plot

pdf('boxplot.pdf', width=14)
boxplot.plot
dev.off()

# Show logFC of genes in pathway
subset(lrt, gene %in% plot.data$gene)
```